---
title: "STAT-731 Recitation 4"
author: "K. Tyler Wilcox"
date: "September 28, 2016"
output:
  beamer_presentation:
    highlight: zenburn
    incremental: false
    theme: "metropolis"
---

Preliminaries
========================================================

```{r, include = FALSE}
knitr::opts_chunk$set(cache = TRUE)
```

- Simulating Random Variables

Discrete Random Variables: Bernoulli Distribution
========================================================

- Consider a random variable $X$, $\mathcal{X} = \{ 0, 1 \}$ with $Pr[X = 1] = \theta$
- The probability mass function of $X$ is:
$$f_X(x) = \theta ^ {x} (1 - \theta) ^ {1 - x}, 0 < \theta  < 1$$
- We used this distribution when we considered a reasonable description of a coin flip

Simulating from the Bernoulli Distribution
========================================================

- Let $\theta = \frac{3}{4}$
- Generate $m = 100$ empirical replications from Ber($\theta$)

```{r}
set.seed(9192016L)
m     = 100L
n     = 1L
theta = 0.75
x     = rbinom(m, n, prob = theta)
```

***

```{r, echo = FALSE, message = FALSE, fig.cap = "Bernoulli(3/4) PMF", fig.height = 4.5}
library(pacman)
p_load(tidyverse, scales)
x = tibble(x = x) %>%
  mutate(x = factor(x))
x %>%
  ggplot() +
  geom_bar(aes(x, ..count.. / sum(..count..))) +
  scale_y_continuous(name = "Relative Frequency", labels = percent,
                     limits = c(0.0, 1.0), breaks = seq(0.0, 1.0, 0.1)) +
  ggtitle("Empirical Probability of X = 1") +
  theme_minimal()
```

***

- Let $\theta = \frac{1}{3}$

```{r}
theta = 1.0 / 3.0
x = rbinom(m, n, prob = theta)
```

```{r, echo = FALSE, message = FALSE, fig.cap = "Bernoulli(1/3) PMF", fig.height = 4.5}
x = tibble(x = x) %>%
  mutate(x = factor(x))
x %>%
  ggplot() +
  geom_bar(aes(x, ..count.. / sum(..count..))) +
  scale_y_continuous(name = "Relative Frequency", labels = percent,
                     limits = c(0.0, 1.0), breaks = seq(0.0, 1.0, 0.1)) +
  ggtitle("Empirical Probability of X = 1") +
  theme_minimal()
```

Discrete Random Variables: Binomial Distribution
========================================================

- Consider $n$ iid Bernoulli random variables $X_1, \ldots, X_n$
- The probability mass function of $X_1, \ldots, X_n$ is:
$$f_X(x) = {n \choose x} \theta ^ {x} (1 - \theta) ^ {n - x}, x \ge 0, \theta > 0$$
- Very common distribution with many applications

Simulating the Binomial Distribution
========================================================

- Let $\theta = \frac{1}{3}$
- Generate $m = 1000$ empirical replications from Bin($n = 3, \theta$)

```{r}
m     = 1000L
n     = 3L
theta = 1.0 / 3.0
x = rbinom(m, n, prob = theta)
```

***

- $n = 3$
- Positively skewed

```{r, echo = FALSE, message = FALSE, fig.cap = "Binomial(n = 3, 1/3) PMF", fig.height = 4.5}
x = tibble(x = x) %>%
  mutate(x = factor(x))
x %>%
  ggplot() +
  geom_bar(aes(x, ..count.. / sum(..count..))) +
  scale_y_continuous(name = "Relative Frequency", labels = percent,
                     limits = c(0.0, 1.0), breaks = seq(0.0, 1.0, 0.1)) +
  ggtitle("Empirical Probability of X") +
  theme_minimal()
```

Gaussian Approximation to Binomial Distribution
========================================================

- If $n$ is large, the Binomial distribution is relatively symmetric
- From the de Moivre-Laplace Theorem:
$$\text{Bin}\left( n, \theta \right) \stackrel{n}{\rightarrow} \text{N}\left(n \theta, n \theta (1 - \theta) \right)$$
- Basis of proportion hypothesis test using $\text{N}(0, 1)$

***

- $n = 5$
- Positively skewed

```{r, echo = FALSE, message = FALSE, fig.cap = "Binomial(n = 5, 1/3) PMF", fig.height = 4.5}
n     = 5L
theta = 1.0 / 3.0
x = rbinom(m, n, prob = theta)
x = tibble(x = x) %>%
  mutate(x = factor(x))
x %>%
  ggplot() +
  geom_bar(aes(x, ..count.. / sum(..count..))) +
  scale_y_continuous(name = "Relative Frequency", labels = percent,
                     limits = c(0.0, 1.0), breaks = seq(0.0, 1.0, 0.1)) +
  ggtitle("Empirical Probability of X") +
  theme_minimal()
```

***

- More symmetric
- Still skewed

```{r, echo = FALSE, message = FALSE, fig.cap = "Binomial(n = 5, 1/3) PMF", fig.height = 4.5}
n     = 10L
theta = 1.0 / 3.0
x = rbinom(m, n, prob = theta)
x = tibble(x = x) %>%
  mutate(x = factor(x))
x %>%
  ggplot() +
  geom_bar(aes(x, ..count.. / sum(..count..))) +
  scale_y_continuous(name = "Relative Frequency", labels = percent,
                     limits = c(0.0, 1.0), breaks = seq(0.0, 1.0, 0.1)) +
  ggtitle("Empirical Probability of X") +
  theme_minimal()
```

***

- $n = 30$
- Symmetric, approaching Gaussian distribution

```{r, echo = FALSE, message = FALSE, fig.cap = "Binomial(n = 5, 1/3) PMF", fig.height = 4.5}
n     = 30L
theta = 1.0 / 3.0
x = rbinom(m, n, prob = theta)
x = tibble(x = x) %>%
  mutate(x = factor(x))
x %>%
  ggplot() +
  geom_bar(aes(x, ..count.. / sum(..count..))) +
  scale_y_continuous(name = "Relative Frequency", labels = percent,
                     limits = c(0.0, 1.0), breaks = seq(0.0, 1.0, 0.1)) +
  ggtitle("Empirical Probability of X") +
  theme_minimal()
```

Wrapping Up
========================================================

- Probability Mass Functions
- Probability Density Functions
- Cumulative Distribution Functions
- Approximation with Monte Carlo
- Questions?
